---
title: "hwk3"
author: "Stephen Carroll"
date: last-modified
execute: 
  eval: true
  warning: false
  message: false
  echo: true
format:
  html:
    toc: true
    code-fold: true
editor_options: 
  chunk_output_type: console
---

```{r}
# Load packages
library(sf)
library(stars)
library(tmap)
library(here)
library(tidyverse)
library(kableExtra)
library(testthat)
library(spData)
library(spDataLarge)
library(geodata)
library(raster)
```

```{r}
# Read in Houston road data
# Exclude data that doesn't include highways
roads <- read_sf(here("data", "data","gis_osm_roads_free_1.gpkg"), query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'", quiet = TRUE)

# Read in Houston house data
# Exclude non-residential buildings
houses <- read_sf(here("data", "data", "gis_osm_buildings_a_free_1.gpkg"), query = "SELECT *
FROM gis_osm_buildings_a_free_1
WHERE (type IS NULL AND name IS NULL)
OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')", quiet = TRUE)

# Read in socioeconomic data
socioeco <- st_read(here("data","data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb"), layer = "ACS_2019_5YR_TRACT_48_TEXAS", quiet = TRUE)

# Read in nightlight data

# Data for 2021-02-07
nightlight_07_05 <- rast(here("data", "data","VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
nightlight_07_06 <- rast(here("data", "data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
  
# Data for 2021-02-16
nightlight_16_05 <- rast(here("data", "data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
nightlight_16_06 <- rast(here("data", "data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```


```{r}
# Make a warning message to ensure CRS is the same across all data frames
check_df_crs <- function() {
  df_crs_comparisons <- list(  # Create a named list with comparisons
    "roads vs houses" = st_crs(roads) == st_crs(houses),
    "roads vs socioeco" = st_crs(roads) == st_crs(socioeco),
    "houses vs socioeco" = st_crs(houses) == st_crs(socioeco),
    "nightlight_07_05 vs nightlight_07_06" = st_crs(nightlight_07_05) == st_crs(nightlight_07_06), 
    "nightlight_07_05 vs nightlight_16_05" = st_crs(nightlight_07_05) == st_crs(nightlight_16_05),
    "nightlight_07_05 vs nightlight_16_06" = st_crs(nightlight_07_05) == st_crs(nightlight_16_06)
  )

# Identify results that are FALSE by comparison name
false_comparisons <- names(df_crs_comparisons)[!unlist(df_crs_comparisons)]

if (length(false_comparisons) == 0) {
    print("All CRS match. ")
} else {
    warning(paste(paste(false_comparisons, "CRS projections do not match.", collapse = " ")))
  }
}

# Call warning function
check_df_crs()
```

```{r}
# Change the CRS of roads to match socioeco and houses
socioeco <- st_transform(socioeco, crs = st_crs(houses))
roads <- st_transform(roads, crs = st_crs(houses))

# Call warning function again to ensure transformation worked
check_df_crs()
```

XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX
XXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXXX

### Find locations that experienced a blackout by creating a mask

```{r}
# Combine the raster data to show all of Houston for the 7th 
lights_07 <- merge(nightlight_07_05, nightlight_07_06)

# Combine the raster data to show all of Houston for the 16th
lights_16 <- merge(nightlight_16_05, nightlight_16_06)


# Find the change in light intensity between the two days
lights_diff <- lights_16 - lights_07
```

```{r}
# Set all cells with values less than 200 to NA
lights_diff[lights_diff < 200] <- NA

```

XXXX
FIND A WAY TO EXPLAIN THIS
XXXX

```{r}
# Use coordinates for the Houston area to make a data frame
coords <- data.frame(x = c(-96.5, -96.5, -94.5, -94.5), 
                     y = c(29, 30.5, 30.5, 29))

# Make the data frame a spatial object 
coordinates(coords) <- ~ x + y

# Change the coords into a polygon object called bbox
bbox <- Polygon(coords)

# Make a spatial polygon that contains the coord bbox
houston_box <- SpatialPolygons(list(Polygons(list(bbox), "bbox")))
```

```{r}
# Crop raster to extent of vector object
lights_cropped <- crop(lights_diff, houston_box)
```

```{r}
# Transform the cropped data to vector objects (polygons)
lights_croppped_vec <- as.polygons(lights_cropped) %>% 
  st_as_sf() %>%
  st_make_valid() 
```

```{r}
# Change the CRS to EPSG:3083
lights_cropped_vec <- st_transform(lights_croppped_vec, crs = "EPSG:3083")
```


XXXXXXXXXXXXXXXXXXXXXXXXX
```{r}
# Create a mask for the 7th
#lights_07_mask <- lights_07

# Create a mask for the 16th
#lights_16_mask <- lights_16

# Create a clip variable with the extent of Houston
#clip <- bbox()

# Clip the data to the extent of Houston
#lights_16_mask_clip <- lights_16_mask[clip]
#lights_07_mask_clip <- lights_07_mask[clip]
```
XXXXXXXXXXXXXXXXXX

XXXXXXXXXXXXXXX
```{r}
tm_shape(lights_croppped_vec) +
  tm_polygons() +
  tm_layout(legend.outside = TRUE, bg.color = "black")
```
XXXXXXXXXXXXXXXXX


Exclude highways from the cropped blackout mask

XXXXXXXXXXXXXXX
MAke a warnign test below for CRS
XXXXXXXXXXXXXXXX
```{r}
# Make a variable for the CRS of the Houston lights dataset
crs_lights_cropped <- st_crs(lights_cropped_vec) 

# Match this CRS variable with the roads dataset
roads <- st_transform(roads, crs = crs_lights_cropped)
```

```{r}
# Make a buffer of 200m to exclude highways and the surrounding areas
roads_buffer_200 <- st_buffer(roads, dist = 200)

# Unionize the polygons into one object and make it valid
roads_buffer_union <- st_union(roads_buffer_200) %>%
  st_make_valid() %>%
  st_as_sf()
```

```{r}
# Find the area that had blackouts that are further than 200m from a highway 
lights_hwy_buffer <- st_difference(lights_cropped_vec, roads_buffer_union)
```

```{r}
# Make a test map that shows the new light map with a highway buffer
lights_hwy_buffer_map <- tm_shape(lights_hwy_buffer) +
  tm_polygons() +
  tm_layout(legend.outside = TRUE, bg.color = "black")
```

```{r}
# Make a test map that shows the light map for the Houston area to compare the two
lights_cropped_map <- tm_shape(lights_cropped_vec) +
  tm_polygons() +
  tm_layout(legend.outside = TRUE, bg.color = "black")

tmap_arrange(lights_cropped_map, lights_hwy_buffer_map, ncol = 2)
```

Identify homes that experienced blackouts by combining the locations of homes and blackouts

```{r}
## Test the CRS here
#st_crs(lights_hwy_buffer) == st_crs(houses)

# Reconcile the CRS of the houses data and the buffered hwy light data
# Make a variable of the CRS for the buffered hwy light data
crs_lights_hwy_buffer <- st_crs(lights_hwy_buffer)

# Set the houses CRS to this variable
houses <- st_transform(houses, crs = crs_lights_hwy_buffer)

```

```{r}
# Make an list of indices of the rows from 'houses' that intersect with the 'lights_hwy_buffer' dataframe
blackout_houses_index <- st_intersects(lights_hwy_buffer, houses)
```

```{r}
# Convert index list to a vector of indices
blackout_houses_index <- blackout_houses_index %>%
  unlist() %>% 
  unique() # Remove duplicate values if present
```

```{r}
# Make a subset dataframe of 'houses' using the corresponding values from the index vector
blackout_houses <- houses[blackout_houses_index, ]
```


```{r}
# Make a map of the houses that were in blackout areas
blackout_houses_map <- tm_shape(blackout_houses) +
  tm_polygons()

blackout_houses_map
```


identify the census tracts likely impacted by blackout


# Plot comparing dist. of median incomes. refer to HW#2 or example